Compositional Certified Resource Bounds\ifshort{\vspace{-1.15cm}}{}

### 2014-11-13 paper/main.tex
some typos

In software engineering and software verification,
we often would like to have static information
about the quantitative behavior of programs.
For example, stack and heap-space bounds
are important to ensure the reliability of
safety-critical systems~\cite{veristack14,Regehr05}.
Static energy usage information is critical
for autonomous systems and has applications in
cloud computing~\cite{CohenZSL12,CarrollH10}.
Worst-case time bounds can help to create
constant-time implementations that prevent
side-channel attacks~\cite{KasperS09,BartheBCLP14}.
Loop and recursion-depth bounds are used to
ensure the accuracy of programs that are executed
on unreliable hardware~\cite{CarbinMR13} and
complexity bounds are needed to verify cryptographic
protocols~\cite{BartheGB09}.  In general, quantitative
resource information at design time can provide useful
feedback for developers.

@@

Available techniques for automatically deriving worst-case resource bounds
fall into two categories.  Techniques
in the first category derive impressive bounds for
numerical imperative programs, but are not fully compositional.
This is problematic if one needs to derive global whole-program bounds.
Techniques in the second category derive tight whole-program
bounds for programs with regular loop and recursion patterns that
decrease the size of an individual variable or data structure.
They are highly compositional, scale for large
programs, and work directly on the syntax.  However, % despite all
% these theoretical and practical advantages,
it has been
a long-time open problem to make them work for typical imperative code
with non-regular iteration patterns, signed integers, mutation, and non-linear
control flow. % As a result, they are not suitable for
% typical imperative code.

Tools in the first category include
SPEED~\cite{GulwaniMC09}, KoAT~\cite{BrockschmidtEFFG14},
PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10},
and LOOPUS~\cite{SinnZV14}.
They lack compositionality in at least two ways.
First, they all base their analysis on some form of
\emph{ranking function} or \emph{counter instrumentation}
that is linked to a local analysis.  As a result, loop bounds are
arithmetic expressions that depend on the values of
variables just before the loop.
This makes it hard to give a
resource bound on a sequence of loops and function calls in
terms of the input parameters of a function.
Second, while all popular imperative programming languages
provide a function or procedure abstraction, available
tools are not able to abstract resource behavior
and have to inline the procedure body to perform their
analysis.

Tools in the second category originate form the potential method of
amortized analysis and type systems for functional
programs~\cite{Jost03,HoffmannAH12}. It has been shown that class
definitions of object oriented programs~\cite{Jost06} and
data-structure predicates of separation logic~\cite{Atkey10} can play
the role of the type system in imperative programs. However, these
techniques fail for numerical imperative programs because these programs lack
fine-grained data-structure information to hook the analysis on.
Another issue is that existing techniques associate \emph{potential}
directly with single program variables.  In imperative programs, this
fails for loops as simple as \lstinline{for(i=x;i<y;i++)} where
$y - i$ decreases, but not $|i|$. %  Mutation is yet another problem.
% While functional programs relate multiple fixed inputs (the typing
% context) with one output (the expression typed), imperative programs
% can change the value of variables in scope, so the potential of these
% variables has to be updated after mutations to ensure soundness.

A general problem with existing tools is user interaction. When a tool
fails to find a resource bound for a program, there is no possibility
for sound user interaction to guide the tool during during bound
derivation.
For example, there is no concept of manual proofs of
resource bounds and no framework that could support composition
of manually derived bounds with automatically inferred bounds.

@@

This paper presents a new framework for automatically
deriving resource bounds on C programs.  This new
approach is an attempt to unify the two previous
categories: It solves the compositionality issues of techniques
for numerical imperative code by adapting amortized-analysis based
techniques from the functional world.
%
Our automated analysis is able to infer
resource bounds on C programs with mutually-recursive functions and integer loops.
%
The resource behavior of functions can be summarized
in a function specification that can be used at every
call-side without accessing the function body.
%
To our knowledge this is the first technique based on
amortized analysis that is able to derive bounds that depend on negative numbers
and differences of variables.  It is also the first
resource analysis technique for C that deals naturally with
recursive functions and sequenced loops, and can handle resources that can
become available during execution (e.g., when freeing
memory).  Compared to more classical
approaches based on ranking functions, our tool inherits
the benefits of amortized reasoning.  Using only one
simple mechanism, it handles:
\itemskip
\begin{itemize}
\item interactions between \emph{sequential loops or
  function calls} through size changes of variables,
\itemskipIn
\item \emph{nested loops} that influence each other
  with the same set of modified variables,
\itemskipIn
\item and \emph{amortized bounds} as found, for example, in
  the Knuth-Morris-Pratt algorithm for string search.
\end{itemize}
\itemskip
The main innovation that makes amortized analysis work
on imperative languages is to base the analysis on a
Hoare-like logic and to track multivariate quantities instead
of program variables.  This leads to precise bounds
expressed as functions of sizes $|[x, y]| = \max(0, y-x)$ of
intervals.
%
A distinctive feature of our analysis system is that
it reduces linear bound inference  to a linear optimization problem
that can solved by off-the-shelf LP solvers.  This enables the
efficient inference of global bounds for larger programs.
%
Moreover, our local inference rules automatically generate
proof certificates that can be easily checked in linear time.

@@

Several micro benchmarks demonstrate the practicality and
expressiveness of the user guided bound inference.  For example, we
derive a logarithmic bound for a binary search function, a bound that
depends on the contents of an array to describe the exact cost of a
function that finds a maximal element in an array, and a bound
that amortizes the cost of $k$ successive increments to a binary
counter (see \pref{sec:anno}).

In summary, we make the following \emph{contributions}.
\itemskip
\begin{itemize}
\item We develop the first automatic amortized analysis for C
  programs. It is naturally compositional, tracks size changes of
  variables to derive global bounds, can handle mutually-recursive
  functions, generates resource abstractions for functions, derives
  proof certificates, and handles resources that can
  become available during execution.
\item We show how to automatically reduce the inference of \emph{linear} resource
    bounds to efficient LP solving.
\item We describe a new method of harnessing existing qualitative
  verification techniques to guide the automatic amortized analysis
  to derive \emph{non-linear} resource bounds with LP solving.
\item We prove the soundness of the analysis with respect to a
  parametric cost semantics for C programs. The cost model can be
  further customized with function calls (\code{tick(n)}) that
  indicate resource usage. % This mechanism is also covered by the
  % soundness proof.
\item We present experiments with a prototype implementation on more
  than 2000 lines of C code. A detailed comparison shows that our
  prototype is the only tool that can derive global bounds for larger
  C programs while being as powerful as existing tools when deriving
  linear local bounds for tricky loop and recursion patterns.
\end{itemize}
\itemskip
%
\iffull{}
{\footnotesize \sl
[To the reviewers: The auxiliary material contains an anonymized
version of the extended article, and the source code of $\toolname$
and all examples.] }

@@

To obtain a compositional reasoning we also have to take into account the
state resulting from a program's execution.  We thus use two potential
functions, one that applies before the execution, and one that applies
after.  The two functions must respect the relation $\Phi(\state)
\ge n + \Phi'(\state')$ for all states $\state$ and $\state'$ such
that $(S, \state) \Downarrow_n \state'$.  Intuitively, $\Phi(\state)$
must provide enough \emph{potential} for both, paying for the resource
cost of the computation and paying for the potential $\Phi'(\state')$ on
the resulting state $\state'$. That way, if $(\state, S_1) \Downarrow_n
\state'$ and $(\state', S_2) \Downarrow_m \state''$, we get $\Phi(\state)
\ge n + \Phi'(\state')$ and $\Phi'(\state') \ge m + \Phi''(\state'')$.
This can be composed as $\Phi(\state) \ge (n + m) + \Phi''(\state'')$.
Note that the initial potential function $\Phi$ provides an upper bound
on the resource consumption of the whole program.  What we have observed
is that, if we define $\htriple{\Phi} {S}{\Phi'}$ to mean
$$
\forall \state\, n\, \state'.\, (\state, S) \Downarrow_n \state' \implies \Phi(\state) \ge
n + \Phi'(\state') \; ,
$$
then we get the following familiar looking rule
$$
\RuleNolabel
{\htriple{\Phi}{S_1}{\Phi'} \\ \htriple{\Phi'}{S_2}{\Phi''}}
{\htriple{\Phi} {S_1; S_2}  {\Phi''}} \;\; .
$$
%
This rule already shows a departure from classical techniques that are
based on ranking functions.  Reasoning with two potential functions
promotes compositional reasoning by considering the sequencing of
program operations.  In the previous rule, $\Phi$ gives a bound for
$S_1; S_2$ through the intermediate potential $\Phi'$, even though it
was derived on $S_1$ only.
%
Similarly, other language constructs lead to rules for the potential
functions that look very similar to Hoare logic or effect system
rules.  These rules enable interactive reasoning about resource usage
in a flexible and compositional way, which, as a side effect, produces
a certificate for the derived resource bound.

@@

\pref{fig:ex1} shows a derivation of the bound
$\frac{T}{K}{\cdot}|[x,y]|$ on the number of ticks for a generalized
version of Example~1 in which we increment $x$ by a constant $K>0$ and
consume $T>0$ resources in each iteration.  The reasoning is similar
to the one of Example~1 except that we obtain the potential
$K{\cdot}\frac{T}{K}$ after the assignment.  Note that the logical
assertions in $\Gamma$ are only used in the rule for the assignment
\code{x=x+K}.

To the best of our knowledge, no other implemented tool for C is
currently capable of deriving a tight bound on the cost of such a
loop.  For $T=1$ (many systems focus on the number of loop iterations
without a cost model) and $K=10$, KoAT computes the bound $|x| + |y| +
10$, Rank computes the bound $y-x-7$, and LOOPUS computes the bound
$y-x-9$.  Only PUBS computes the tight bound $0.1(y-x)$ if we
translate the program into a term-rewriting system by
hand. Nevertheless, we will show in \pref{sec:AAA} that the
potential method makes automatic bound derivation straightforward.

@@

The concept of a potential function is a generalization of the
concept of a ranking function.  A potential function can be used like
a ranking function if we use the tick metric and add the statement
\code{tick(1)} to every back edge of the program (loops and function
calls).  However, a potential function is more flexible.  For example,
we can use a potential function to prove that Example~2 does not
consume any resources in the tick metric.
\begin{lstlisting}[basicstyle=\tt\small]
#\!\!\!\!\!# while (x<y) {tick(-1); x=x+1; tick(1)} #\hfill \textnormal{(Example 2)}#
\end{lstlisting}
Similarly we can prove that Example~3 can be bounded by $10|[x,y]|$.
In both cases, we reason exactly like in the first version of the
while loop to prove the bound.  Of course, such loops with different
tick annotations can be seamlessly combined in a larger program.
\begin{lstlisting}[basicstyle=\tt\small]
while (x<y) { x=x+1; tick(10); } #\hfill \textnormal{(Example 3)}#
\end{lstlisting}

@@

To find resource bounds automatically, we first need to restrict our
search space.  In this work, we focus on a the following form of
potential functions, which can express tight bounds for many typical
programs and allows inference view \emph{linear programming}.
$$
  \Phi(\state) = q_0 + \sum_{x, y \in \dom\state \land x \neq y}
    q_{(x, y)} \cdot |\inter {\state(x)} {\state(y)}| \, .
$$
%
Here $\state : (\text{Locals} \to \Z) \times (\text{Globals} \to \Z)$
is a simplified program state that maps variable names to integers,
$|\inter a b| = \max(0,b-a)$, and $q_i \in \Qplusz$.  To simplify the
references to the linear coefficients $q_i$, we introduce an \emph{index
  set} $I$.  This set is defined to be $\{0\} \cup \{(x, y) \mid x, y
\in \Var \land x \neq y \}$.  Each index $i$ corresponds to a \emph{base function}
$f_i$ in the potential function: $0$ corresponds to the constant
function $\state \mapsto 1$, and $(x,y)$ corresponds to $\state \mapsto
|\inter {\state(x)} {\state(y)}|$.  Using these notations we can
rewrite the above equality as\iffull{
$$ \Phi(\state) = \sum_{i \in I} q_i f_i(\state).$$}{
$\Phi(\state) = \sum_{i \in I} q_i f_i(\state)$.}%
We often write $xy$ to denote the
index $(x,y)$.  \iffull{The family $(f_i)_{i\in I}$ is actually a basis (in
the linear-algebra sense) for potential functions.}{}  This allows us to
uniquely represent any linear potential function
$\Phi$ as a \emph{quantitative annotation} $Q = (q_i)_{i \in I}$, that
is, a family of \ifshort{non-negative}{} rational numbers where only
a finite number of elements are not zero. \iffull{ For soundness, we
  require that potential functions always give non-negative results.
  This is achieved by restricting coefficients in a quantitative
  annotation to be always non-negative.}{}

@@

In contrast to the quantitative annotations, logical contexts $\Gamma$
in the presented inference system are left abstract.  This allows for
simpler rules and leaves room for future improvements.
Our implementation finds these logical contexts using abstract
interpretation using the domain of linear inequalities.  We observed that
the rules of the analysis often requires only minimal local knowledge.
This means that it is not necessary for us to compute precise loop
invariants and only a rough fixpoint (e.g. keeping only inequalities on
variables unchanged by the loop) is sufficient to obtain good bounds.

@@

   \end{lstlisting}
\vspace{-2.5ex}
\begin{lstlisting}
for (;;) {
  do { l++; tick(1); }
    while (l<h && *);
  do { h--; tick(1); }
    while (h>l && *);
  if (h<=l) break;
  tick(1); /* swap elems. */ }
\end{lstlisting}
$2+3|[l,h]|$
\\[.4\baselineskip]
      {\bf t62}
    \end{center}
  \end{minipage}
\vspace{1ex}
\caption{Example \emph{t39} shows two mutually-recursive functions
  with the computed tick bounds.  Example \emph{t61} and \emph{t62}
  demonstrate the unique compositionality of our system. In \emph{61},
  $N\geq 0$ is a fixed but arbitrary constant.  The derived constant factors
  are tight.}
  \label{fig:ex_list_2}
\end{figure*}

@@

One might think that our set of potential functions is too simplistic
to be able to express and prove bounds for realistic
programs. Nevertheless, we can handle challenging example programs
without special tricks or techniques.  Examples \emph{speed\_1} and
\emph{speed\_2}, that are taken from previous work~\cite{GulwaniMC09},
demonstrate that our method can handle \emph{tricky iteration
  patterns}.  The SPEED tool~\cite{GulwaniMC09} derives the same
bounds as our analysis but requires heuristics for its counter
instrumentation.  These loops can also be handled with inference of
\emph{disjunctive invariants}, but in the abstract interpretation
community, these invariants are known to be notoriously difficult to
generate.
%
In example \emph{speed\_1} we have one loop that first increments
variable $y$ up to $m$ and then increments variable $x$ up to $n$.  We
derive the tight bound $|[x, n]| + |[y, m]|$.
%
Example \emph{speed\_2} is even trickier, and we found it hard to
find a bound manually.  However, using potential transfer reasoning as
in amortized analysis, it is easy to prove the tight bound
$|[x, n]| + |[z, n]|$.

@@

Example \emph{t08a} shows the ability of the analysis to discover
interaction between \emph{sequenced loops} through size change of
variables.  We accurately track the size change of $y$ in the first
loop by transferring the potential $0.1$ from $|[y,z]|$ to $|[0,y]|$.
Furthermore, \emph{t08a} shows again that we do not handle the
constants $1$ or $0$ in any special way.  In all examples we could
replace $0$ and $1$ with other constants like in the second loop and
still derive a tight bound.  \iffull{The only information, that the analyzer
needs is $y \geq c$ before assigning $y = y - c$.}{}
%
Example \emph{t27} shows how amortization can be used to handle
\emph{interacting nested loops}.  In the outer loop we increment the
variable $n$ until $n = 0$.  In each of the $|[n,0]|$ iterations, we
increment the variable $y$ by $1000$.  Then we non-deterministically
(expressed by \code{*}) execute an inner loop that decrements $y$ by
$100$ until $y<100$.  The analysis discovers that only the first
execution of the inner loop depends on the initial value of $y$.  We
again derive tight constant factors.
%

@@

Example \emph{t61} in \pref{fig:ex_list_2} is typical for implementations
of block-based cryptographic primitives: Data of arbitrary length, is
consumed in blocks and the leftover is stored in a buffer for future use
when more data is available.  It is present in all the block encryption
routines of PGP.  The same pattern is also used in performance critical
code to unroll a loop, for example we found it in a bit manipulating
function of libtiff library and a CRC computation routine of MAD,
an MPEG audio decoder.  This looping pattern is particularly well
handled by our method. If $N \geq 8$, $\toolname$ infers the bound
$\frac{N}{8} |[0,l]|$, but if $N<8$ $\toolname$ infers $7\frac{8-N}{8} +
\frac{N}{8}|[0,l]|$. The selection of the block size ($8$) and the cost
in the second loop (\code{tick(1)}) are random choices and $\toolname$
would also derive tight bound for other values.

@@

The derivation system for the automatic amortized analysis is defined
in~\pref{fig:auto}.  The derivation rules derive judgements of the form
$$
  (\Gamma_B, Q_B); (\Gamma_R, Q_R);
    (\Gamma, Q) \vdash S \dashv (\Gamma', Q').
$$
%
The part $(\Gamma, Q) \vdash S \dashv (\Gamma', Q')$ of the judgement
can be seen as a regular Hoare triple.  All assertions are split into
two parts, the logical part and the quantitative part.  The
quantitative part $Q$ represents a potential function as a collection
of non-negative numbers $q_i$ indexed by the index set $I$.  The
logical part $\Gamma$ is left abstract but is enforced by our
derivation system to respect classic Hoare logic constraints.
%
The meaning of this basic judgment is as follows: If $S$ is executed
with starting state $\state$, the assertions in $\Gamma$
hold, and at least $Q(\state)$ resource units are available then the
evaluation does not run out of resources and---if the execution
terminates in state $\state'$---there are at least $Q(\state')$
resources left and $\Gamma'$ holds for $\state'$.

>>>>

In software engineering and software verification,
we often would like to have static information
about the quantitative behavior of programs.
For example, stack and heap-space bounds
are important to ensure the reliability of
safety-critical systems~\cite{Regehr05}.
Static energy usage information is critical
for autonomous systems and has applications in
cloud computing~\cite{CohenZSL12,CarrollH10}.
Worst-case time bounds can help create
constant-time implementations that prevent
side-channel attacks~\cite{KasperS09,BartheBCLP14}.
Loop and recursion-depth bounds are used to
ensure the accuracy of programs that are executed
on unreliable hardware~\cite{CarbinMR13} and
complexity bounds are needed to verify cryptographic
protocols~\cite{BartheGB09}.  In general, quantitative
resource information at design time can provide useful
feedback for developers.

@@

Available techniques for automatically deriving worst-case resource bounds
fall into two categories.  Techniques
in the first category derive impressive bounds for
numerical imperative programs, but are not compositional.
This is problematic if one needs to derive global whole-program bounds.
Techniques in the second category derive tight whole-program
bounds for programs with regular loop and recursion patterns that
decrease the size of an individual variable or data structure.
They are highly compositional, scale for large
programs, and work directly on the syntax.  However, % despite all
% these theoretical and practical advantages,
it has been
a long-time open problem to make them work for typical imperative code
with non-regular iteration patterns, signed integers, mutation, and non-linear
control flow. % As a result, they are not suitable for
% typical imperative code.

Tools in the first category include
SPEED~\cite{GulwaniMC09}, KoAT~\cite{BrockschmidtEFFG14},
PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10},
and LOOPUS~\cite{SinnZV14}.
They lack compositionality in at least two ways.
First, they all base their analysis on some form of
\emph{ranking function} or \emph{counter instrumentation}
that is linked to a local analysis.  As a result, loop bounds are
arithmetic expressions that depend on the values of
variables just before the loop.
This makes it hard to give a
resource bound on a sequence of loops and function calls in
terms of the input parameters of a function.
Second, while all popular imperative programming languages
provide a function or procedure abstraction, available
tools are not able to abstract resource behavior;
instead, they have to inline the procedure body to perform their
analysis.

Tools in the second category originate form the potential method of
amortized analysis and type systems for functional
programs~\cite{Jost03,HoffmannAH12}. It has been shown that class
definitions of object oriented programs~\cite{Jost06} and
data-structure predicates of separation logic~\cite{Atkey10} can play
the role of the type system in imperative programs. However, these
techniques fail for numerical imperative programs because they lack
fine-grained data-structure information to hook the analysis on.
Another issue is that existing techniques associate \emph{potential}
directly with single program variables.  In imperative programs, this
fails for loops as simple as \lstinline{for(i=x;i<y;i++)} where
$y - i$ decreases, but not $|i|$. %  Mutation is yet another problem.
% While functional programs relate multiple fixed inputs (the typing
% context) with one output (the expression typed), imperative programs
% can change the value of variables in scope, so the potential of these
% variables has to be updated after mutations to ensure soundness.

A general problem with existing tools is user interaction. When a tool
fails to find a resource bound for a program, there is no possibility
for sound user interaction to guide the tool during bound
derivation.
For example, there is no concept of manual proofs of
resource bounds and no framework can support composition
of manually derived bounds with automatically inferred bounds.

@@

This paper presents a new framework for automatically
deriving resource bounds on C programs.  This new
approach is an attempt to unify the two aforementioned
categories: It solves the compositionality issues of techniques
for numerical imperative code by adapting amortized-analysis--based
techniques from the functional world.
%
Our automated analysis is able to infer
resource bounds on C programs with mutually-recursive functions and integer loops.
%
The resource behavior of functions can be summarized
in a function specification that can be used at every
call site without accessing the function body.
%
To our knowledge this is the first technique based on
amortized analysis that is able to derive bounds that depend on negative numbers
and differences of variables.  It is also the first
resource analysis technique for C that deals naturally with
recursive functions and sequenced loops, and can handle resources that can
become available during execution (e.g., when freeing
memory).  Compared to more classical
approaches based on ranking functions, our tool inherits
the benefits of amortized reasoning.  Using only one
simple mechanism, it handles:
\itemskip
\begin{itemize}
\item interactions between \emph{sequential loops or
  function calls} through size changes of variables,
\itemskipIn
\item \emph{nested loops} that influence each other
  with the same set of modified variables,
\itemskipIn
\item and \emph{amortized bounds} as found, for example, in
  the Knuth-Morris-Pratt algorithm for string search.
\end{itemize}
\itemskip
The main innovation that makes amortized analysis work
on imperative languages is to base the analysis on a
Hoare-like logic and to track multivariate quantities instead
of program variables.  This leads to precise bounds
expressed as functions of sizes $|[x, y]| = \max(0, y-x)$ of
intervals.
%
A distinctive feature of our analysis system is that
it reduces linear bound inference  to a linear optimization problem
that can solved by off-the-shelf LP solvers.  This enables the
efficient inference of global bounds for larger programs.
%
Moreover, our local inference rules automatically generate
proof certificates that can be easily checked in linear time.

@@

Examples from cBench and micro benchmarks demonstrate the practicality and
expressiveness of the user guided bound inference.  For example, we
derive a logarithmic bound for a binary search function\iffull{, a bound that
depends on the contents of an array to describe the exact cost of a
function that finds a maximal element in an array, }{ }and a bound
that amortizes the cost of $k$\iffull{ successive }{ }increments to a binary
counter (see \pref{sec:anno}).

In summary, we make the following \emph{contributions}.
\itemskip
\begin{itemize}
\item We develop the first automatic amortized analysis for C
  programs. It is naturally compositional, tracks size changes of
  variables to derive global bounds, can handle mutually-recursive
  functions, generates resource abstractions for functions, derives
  proof certificates, and handles resources that can
  become available during execution.
\itemskipIn
\item We show how to automatically reduce the inference of \emph{linear} resource
    bounds to efficient LP solving.
\itemskipIn
\item We describe a new method of harnessing existing qualitative
  verification techniques to guide the automatic amortized analysis
  to derive \emph{non-linear} resource bounds with LP solving.
\itemskipIn
\item We prove the soundness of the analysis with respect to a
  parametric cost semantics for C programs. The cost model can be
  further customized with function calls (\code{tick(n)}) that
  indicate resource usage. % This mechanism is also covered by the
  % soundness proof.
\itemskipIn
\item We implemented our resource bound analysis in the publically-available tool $\toolname$.
\itemskipIn
\item We present experiments with $\toolname$ on more
  than 2000 lines of C code. A detailed comparison shows that our
  prototype is the only tool that can derive global bounds for larger
  C programs while being as powerful as existing tools when deriving
  linear local bounds for tricky loop and recursion patterns.
\end{itemize}
\itemskip
%
\iffull{}
{\footnotesize \sl
[To the reviewers: The auxiliary material contains an anonymized
version of the extended article, and the source code of $\toolname$
and all examples.] }

@@

To obtain a compositional reasoning we also have to take into account the
state resulting from a program's execution.  We thus use two potential
functions, one that applies before the execution, and one that applies
after.  The two functions must respect the relation $\Phi(\state)
\ge n + \Phi'(\state')$ for all states $\state$ and $\state'$ such
that $(S, \state) \Downarrow_n \state'$.  Intuitively, $\Phi(\state)$
must provide enough \emph{potential} for both, paying for the resource
cost of the computation and paying for the potential $\Phi'(\state')$ on
the resulting state $\state'$. That way, if $(\state, S_1) \Downarrow_n
\state'$ and $(\state', S_2) \Downarrow_m \state''$, we get $\Phi(\state)
\ge n + \Phi'(\state')$ and $\Phi'(\state') \ge m + \Phi''(\state'')$.
This can be composed as $\Phi(\state) \ge (n + m) + \Phi''(\state'')$.
Note that the initial potential function $\Phi$ provides an upper bound
on the resource consumption of the whole program.  What we have observed
is that, if we define $\htriple{\Phi} {S}{\Phi'}$ to mean
$$
\forall \state\, n\, \state'.\, (\state, S) \Downarrow_n \state' \implies \Phi(\state) \ge
n + \Phi'(\state') \; ,
$$
then we get the following familiar looking rule
$$
\RuleNolabel
{\htriple{\Phi}{S_1}{\Phi'} \\ \htriple{\Phi'}{S_2}{\Phi''}}
{\htriple{\Phi} {S_1; S_2}  {\Phi''}} \;\; .
$$
%
This rule already shows a departure from classical techniques that are
based on ranking functions.  Reasoning with two potential functions
promotes compositional reasoning by focusing on the sequencing of
programs.  In the previous rule, $\Phi$ gives a bound for
$S_1; S_2$ through the intermediate potential $\Phi'$, even though it
was derived on $S_1$ only.
%
Similarly, other language constructs lead to rules for the potential
functions that look very similar to Hoare logic or effect system
rules.  These rules enable reasoning about resource usage
in a flexible and compositional way, which, as a side effect, produces
a certificate for the derived resource bound.

@@

\pref{fig:ex1} shows a derivation of the bound
$\frac{T}{K}{\cdot}|[x,y]|$ on the number of ticks for a generalized
version of Example~1 in which we increment $x$ by a constant $K>0$ and
consume $T>0$ resources in each iteration.  The reasoning is similar
to the one of Example~1 except that we obtain the potential
$K{\cdot}\frac{T}{K}$ after the assignment.  Note that the logical
assertions are only used in the rule for the assignment
\code{x=x+K}.

To the best of our knowledge, no other implemented tool for C is
currently capable of deriving a tight bound on the cost of such a
loop.  For $T=1$ (many systems focus on the number of loop iterations
without a cost model) and $K=10$, KoAT computes the bound $|x| + |y| +
10$, Rank computes the bound $y-x-7$, and LOOPUS computes the bound
$y-x-9$.  Only PUBS computes the tight bound $0.1(y-x)$ if we
translate the program into a term-rewriting system by
hand. We will show in the following sections that the
potential method makes automatic bound derivation straightforward.

@@

The concept of a potential function is a generalization of the
concept of a ranking function.  A potential function can be used like
a ranking function if we use the tick metric and add the statement
\code{tick(1)} to every back edge of the program (loops and function
calls).  However, a potential function is more flexible.  For example,
we can use a potential function to prove that Example~2 does not
consume any resources in the tick metric.
\begin{lstlisting}[basicstyle=\tt\small]
#\!\!\!\!\!#while (x<y) {tick(-1); x=x+1; tick(1)} #\hfill \textnormal{(Example 2)}#

#\!\!\!\!\!#while (x<y) { x=x+1; tick(10); } #\hfill \textnormal{(Example 3)}#
\end{lstlisting}
Similarly we can prove that Example~3 can be bounded by $10|[x,y]|$.
In both cases, we reason exactly like in the first version of the
while loop to prove the bound.  Of course, such loops with different
tick annotations can be seamlessly combined in a larger program.

@@

To find resource bounds automatically, we first need to restrict our
search space.  In this work, we focus on a the following form of
potential functions, which can express tight bounds for many typical
programs and allows for inference with \emph{linear programming}.
$$
  \Phi(\state) = q_0 + \sum_{x, y \in \dom\state \land x \neq y}
    q_{(x, y)} \cdot |\inter {\state(x)} {\state(y)}| \, .
$$
%
Here $\state : (\text{Locals} \to \Z) \times (\text{Globals} \to \Z)$
is a simplified program state that maps variable names to integers,
$|\inter a b| = \max(0,b-a)$, and $q_i \in \Qplusz$.  To simplify the
references to the linear coefficients $q_i$, we introduce an \emph{index
  set} $I$.  This set is defined to be $\{0\} \cup \{(x, y) \mid x, y
\in \Var \land x \neq y \}$.  Each index $i$ corresponds to a \emph{base function}
$f_i$ in the potential function: $0$ corresponds to the constant
function $\state \mapsto 1$, and $(x,y)$ corresponds to $\state \mapsto
|\inter {\state(x)} {\state(y)}|$.  Using these notations we can
rewrite the above equality as\iffull{
$$ \Phi(\state) = \sum_{i \in I} q_i f_i(\state).$$}{
$\Phi(\state) = \sum_{i \in I} q_i f_i(\state)$.}%
We often write $xy$ to denote the
index $(x,y)$.  \iffull{The family $(f_i)_{i\in I}$ is actually a basis (in
the linear-algebra sense) for potential functions.}{}  This allows us to
uniquely represent any linear potential function
$\Phi$ as a \emph{quantitative annotation} $Q = (q_i)_{i \in I}$, that
is, a family of \ifshort{non-negative}{} rational numbers where only
a finite number of elements are not zero. \iffull{ For soundness, we
  require that potential functions always give non-negative results.
  This is achieved by restricting coefficients in a quantitative
  annotation to be always non-negative.}{}

@@

In contrast to the quantitative annotations, logical contexts $\Gamma$
in the presented inference system are left abstract.  This allows for
simpler rules and leaves room for future improvements.
Our implementation finds these logical contexts using abstract
interpretation with the domain of linear inequalities.  We observed that
the rules of the analysis often requires only minimal local knowledge.
This means that it is not necessary for us to compute precise loop
invariants and only a rough fixpoint (e.g. keeping only inequalities on
variables unchanged by the loop) is sufficient to obtain good bounds.

@@

   \end{lstlisting}
\vspace{-2.5ex}
\begin{lstlisting}
for (;;) {
  do { l++; tick(1); }
    while (l<h && *);
  do { h--; tick(1); }
    while (h>l && *);
  if (h<=l) break;
  tick(1); /* swap elems. */ }
\end{lstlisting}
$2+3|[l,h]|$
\\[.4\baselineskip]
      {\bf t62}
    \end{center}
  \end{minipage}
\vspace{1ex}
\caption{Example \emph{t39} shows two mutually-recursive functions
  with the computed tick bounds.  Example \emph{t61} and \emph{t62}
  demonstrate the unique compositionality of our system. In \emph{t61},
  $N\geq 0$ is a fixed but arbitrary constant.  The derived constant factors
  are tight.}
  \label{fig:ex_list_2}
\end{figure*}

@@

One might think that our set of potential functions is too simplistic
to be able to express and prove bounds for realistic
programs. Nevertheless, we can handle challenging example programs
without special tricks or techniques.  Examples \emph{speed\_1} and
\emph{speed\_2} in Figure~\ref{fig:ex_list}, which are taken from previous work~\cite{GulwaniMC09},
demonstrate that our method can handle \emph{tricky iteration
  patterns}.  The SPEED tool~\cite{GulwaniMC09} derives the same
bounds as our analysis but requires heuristics for its counter
instrumentation.  These loops can also be handled with inference of
\emph{disjunctive invariants}, but in the abstract interpretation
community, these invariants are known to be notoriously difficult to
generate.
%
In example \emph{speed\_1} we have one loop that first increments
variable $y$ up to $m$ and then increments variable $x$ up to $n$.  We
derive the tight bound $|[x, n]| + |[y, m]|$.
%
Example \emph{speed\_2} is even trickier, and we found it hard to
find a bound manually.  However, using potential transfer reasoning as
in amortized analysis, it is easy to prove the tight bound
$|[x, n]| + |[z, n]|$.

@@

Example \emph{t08a} in Figure~\ref{fig:ex_list} shows the ability of the analysis to discover
interaction between \emph{sequenced loops} through size change of
variables.  We accurately track the size change of $y$ in the first
loop by transferring the potential $0.1$ from $|[y,z]|$ to $|[0,y]|$.
Furthermore, \emph{t08a} shows again that we do not handle the
constants $1$ or $0$ in any special way.  In all examples we could
replace $0$ and $1$ with other constants like in the second loop and
still derive a tight bound.  \iffull{The only information, that the analyzer
needs is $y \geq c$ before assigning $y = y - c$.}{}
%
Example \emph{t27} in Figure~\ref{fig:ex_list}
shows how amortization can be used to handle
\emph{interacting nested loops}.  In the outer loop we increment the
variable $n$ until $n = 0$.  In each of the $|[n,0]|$ iterations, we
increment the variable $y$ by $1000$.  Then we non-deterministically
(expressed by \code{*}) execute an inner loop that decrements $y$ by
$100$ until $y<100$.  The analysis discovers that only the first
execution of the inner loop depends on the initial value of $y$.  We
again derive tight constant factors.
%

@@

Example \emph{t61} in \pref{fig:ex_list_2} is typical for implementations
of block-based cryptographic primitives: Data of arbitrary length, is
consumed in blocks and the leftover is stored in a buffer for future use
when more data is available.  It is present in all the block encryption
routines of PGP.  The same pattern is also used in performance critical
code to unroll a loop, for example we found it in a bit manipulating
function of libtiff library and a CRC computation routine of MAD,
an MPEG audio decoder.  This looping pattern is handled particularly well
by our method. If $N \geq 8$, $\toolname$ infers the bound
$\frac{N}{8} |[0,l]|$, but if $N<8$, $\toolname$ infers $7\frac{8-N}{8} +
\frac{N}{8}|[0,l]|$. The selection of the block size ($8$) and the cost
in the second loop (\code{tick(1)}) are random choices and $\toolname$
would also derive tight bound for other values.

@@

The derivation system for the automatic amortized analysis is defined
in~\pref{fig:auto}.  The derivation rules derive judgements of the form
$$
  (\Gamma_B, Q_B); (\Gamma_R, Q_R);
    (\Gamma, Q) \vdash S \dashv (\Gamma', Q').
$$
%
The part $(\Gamma, Q) \vdash S \dashv (\Gamma', Q')$ of the judgement
can be seen as a quantitative Hoare triple.  All assertions are split into
two parts, the logical part and the quantitative part.  The
quantitative part $Q$ represents a potential function as a collection
of non-negative numbers $q_i$ indexed by the index set $I$.  The
logical part $\Gamma$ is left abstract but is enforced by our
derivation system to respect classic Hoare logic constraints.
%
The meaning of this basic judgment is as follows: If $S$ is executed
with starting state $\state$, the assertions in $\Gamma$
hold, and at least $Q(\state)$ resource units are available then the
evaluation does not run out of resources and---if the execution
terminates in state $\state'$---there are at least $Q(\state')$
resources left and $\Gamma'$ holds for $\state'$.
